Deployable Probabilistic Programming

### 2019-04-19 paper/paper.tex
restructured slightly
plan for autodiff
autodiff in progress

\section{Model Syntax and Rules of Differentiation}

@@

\section{Inference}

@@

\subsection{Composability of Models}

>>>>

\subsection{Model Syntax and Rules of Differentiation}

@@

\subsection{Inference}

@@

Infergo employs reverse-mode autodifferentiation via source code
transformation~\cite{GW08}. The reverse mode is the default
choice in machine learning applications, because it works best
with differentiating a scalar return value with respect to
multiple parameters. Automatic differentiation is usually
implemented using either operator overloading or source code
transformation. Operator overloading is often the method of
choice~\cite{Stan17,GXG18,PGC+17} due to relative simplicity,
but the differentiated code must use a special type (on which
the operators are overloaded) for representing floating point
numbers. Regardless of the issue of computational overhead, this
restricts the language used to specify the model to operations on
that special type and impairs interoperability: library
functions cannot be directly called but have to use an adaptor,
or a `lifted' version of the standard library must be supplied.
Similarly, the data must be cast into the special type before it
is passed to the model for inference.

Source code transformation allows to automatically differentiate
program code operating on the built-in floating point type. This
method involves generation of new code, for the function
gradient or for the function itself, ahead of time or on the
fly. To achieve the best performance, some implementations
generate static code for computing the function
gradient~\cite{MMW18,I18}. This requires a rather elaborated
source code analysis, and although a significant progress has
been made in this direction, not all language constructs are
supported. Alternatively, the code of the function itself can be
transformed by augmenting arithmetics and function calls on the
floating point type by function calls that record the
computations on the \textit{tape}, just like in the case of
operator overloading. Then, the backward pass is identical to
that of operator overloading. This method allows the
use of the built-in floating point type along with
straightforward virtually unrestricted language for model
specification, at the cost of slight performance loss.

This last method is used in Infergo: the model is written in Go
and is placed in a separate Go package. A command-line utility
is run on the package to generate a sub-package containing a
differentiated version of the model.  Both the original and the
differentiated model are available simultaneously in the calling
code, so that invocations of model methods which do not require
computation of gradients can be performed without the overhead
of automatic differentiation.  However, the performance loss is
rarely a major issue --- differentiated versions of model
methods are only 2--4 times slower on models from the case
studies~\ref{sec:case-studies}, and the backward pass, which
computes the gradient, is roughly as fast as the original,
undifferentiated version. At least partially, fast backward pass
is due to careful implementation and thorough profiling of
the tape.

% Elementals. Differentiation
%of calls. Distributions. Composability. Nested inference.
